Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-30 main.tex
Add details to exponent idea

\subsection{Other Idea}
There are two main bottlenecks when casting MAP as an iterative algorithm. 
\begin{enumerate}
	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
	\item The second one is the step-size scheduling. $O(\inv{t})$ is decreasing too fast for most proof techniques. Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
\end{enumerate}
Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size. Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
\end{align}
where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
On the other hand, the problem of most Bregman analysis is that they do not get this $\lr^2$ in front the variance. Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
\end{align}
The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size. Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
Now our question is what happens if we get something in-between ? e.g.
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
\end{align}
where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.

@@

\paragraph{Descent Lemma and Telescopic Sum.}
Similarly, applying the three points lemma with $\natp^*$, and taking te expectation with respect to $X_{t+1}$ yields
\begin{align}
	\lr_t(\E[f(\natp_{t+1},X_{t+1})] - f(\natp^*)) +  \E[\cB(\natp_{t+1} ; \natp_{t})] 
	&\leq  \cB( \natp^*; \natp_{t})  - \E[\cB( \natp^*; \natp_{t+1})] \; .
\end{align}

taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_{t+1}) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail. This may be suited for online learning, but not for our stationary setting.

\paragraph{Application to the MAP.}
The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g. exactly the same assumption as in \citet{hanzely2018fastest}. However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which is an average in the primal space, and gives more weight to the first iterates. It does not immediately apply to the last iterate.

@@

\subsection{Work by Loizou}
An assumption pointed out by Nicolas Loizou is 
\begin{align}
	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
\end{align}
for which a sufficient condition is 
\begin{align}
\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
\end{align}
which holds in a lot of settings but not all. For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac. However we can overcome this issue by grouping data points : a gaussian can not overfit on two points. Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$. 
So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)

>>>>

\subsection{Other Idea}
There are two main bottlenecks when casting MAP as an iterative algorithm. 
\begin{enumerate}
	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
	\item The second one is the step-size scheduling. $O(\inv{t})$ is decreasing too fast for most proof techniques. Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
\end{enumerate}
Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} or \citet{dragomir2021fast} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size. Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
\end{align}
where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
On the other hand, the problem of most Bregman analysis \citep{hanzely2018fastest,dorazio2021stochastic} is that they do not get this $\lr^2$ in front the variance. Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
\end{align}
The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size. Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
Now our question is what happens if we get something in-between ? e.g.
\begin{align}
	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
\end{align}
where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.

@@

\paragraph{Descent Lemma and Telescopic Sum.}
Similarly, applying the three points lemma with $\natp^*$ yields
\begin{align}
	\lr_t(f(\natp_{t+1},X_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t})
	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \; .
\end{align}
Ignoring the distance between iterates $\cB(\natp_{t+1} ; \natp_{t})
$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_{t+1}) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail. This may be suited for online learning, but not for our stationary setting.

\paragraph{Application to the MAP.}
The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g. exactly the same assumption as in \citet{hanzely2018fastest}.

@@

\subsection{Work by Loizou}
An assumption pointed out by Nicolas Loizou is 
\begin{align}
	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
\end{align}
for which a sufficient condition is 
\begin{align}
\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
\end{align}
which holds in a lot of settings but not all. For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac. However we can overcome this issue by grouping data points : a gaussian can not overfit on two points. Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$. 
So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$.
As mentioned earlier, though, this assumption does not convert to a $O(t^{-1})$ rate when the step-size is itself $O(t^{-1})$.
